---
title: "Quality control"
author: "Jordi Camps"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: html_document
---

```{r setup, include=FALSE}
library(scater)
library(knitr)
library(SingleCellExperiment)
library(magrittr)
options(stringsAsFactors = FALSE)
```

# Prepare dataset
## Load counts and annotation file
```{r}
reads <- read.csv("Data/reads_mtDNA.csv")
rownames(reads) <- reads[, 1]
reads <- reads[, -1]
anno <- read.csv("Data/annotation.csv", header = TRUE)
anno$sample_id <- gsub("-", ".", anno$sample_id) #match sample_id with columns reads
anno <- anno[order(anno$sample_id),] #order according to sample_id
```

### Inspect expression matrix
```{r}
knitr::kable(
    head(reads[ , 1:3]), booktabs = TRUE,
    caption = 'A table of the first 6 rows and 3 columns of the molecules table.'
)
```

### Inspect annotation matrix
```{r}
knitr::kable(
    head(anno), booktabs = TRUE,
    caption = 'A table of the first 6 rows of the anno table.'
)
```

### Create SingleCellExperiment object
```{r}
reads <- SingleCellExperiment(assays = list(counts = as.matrix(reads)), 
                              colData = anno)
```

### Remove genes
```{r}
keep_feature <- rowSums(counts(reads) > 0) > 0
reads <- reads[keep_feature, ]
```

### Define mtDNA and ERCC
```{r}
isSpike(reads, "ERCC") <- grepl("^ERCC-", rownames(reads))
isSpike(reads, "MT") <- rownames(reads) %in% 
  c("ENSMUSG00000064336", "ENSMUSG00000064337", "ENSMUSG00000064338",
    "ENSMUSG00000064339", "ENSMUSG00000064340", "ENSMUSG00000064341",
    "ENSMUSG00000064342", "ENSMUSG00000064343", "ENSMUSG00000064344",
    "ENSMUSG00000064345", "ENSMUSG00000064346", "ENSMUSG00000064347", 
    "ENSMUSG00000064348", "ENSMUSG00000064349", "ENSMUSG00000064350",
    "ENSMUSG00000064351", "ENSMUSG00000064352", "ENSMUSG00000064353",
    "ENSMUSG00000064354", "ENSMUSG00000064355", "ENSMUSG00000064356",
    "ENSMUSG00000064357", "ENSMUSG00000064358", "ENSMUSG00000064359",
    "ENSMUSG00000064360", "ENSMUSG00000064361", "ENSMUSG00000065947",
    "ENSMUSG00000064363", "ENSMUSG00000064364", "ENSMUSG00000064365",
    "ENSMUSG00000064366", "ENSMUSG00000064367", "ENSMUSG00000064368",
    "ENSMUSG00000064369", "ENSMUSG00000064370", "ENSMUSG00000064371",
    "ENSMUSG00000064372")

reads <- calculateQCMetrics(reads,
  feature_controls = list(ERCC = isSpike(reads, "ERCC"), MT = isSpike(reads, "MT"))
)
```

### Annotate genes
```{r}
reads <- getBMFeatureAnnos(
  reads,
  filters = "ensembl_gene_id",
  attributes = c(
    "ensembl_gene_id",
    "mgi_symbol",
    "chromosome_name",
    "transcript_biotype",
    "transcript_start",
    "transcript_end",
    "transcript_count"),
  feature_symbol = "mgi_symbol",
  feature_id = "ensembl_gene_id",
  biomart = "ENSEMBL_MART_ENSEMBL",
  dataset = "mmusculus_gene_ensembl",
  host = "www.ensembl.org"
)
rownames(reads) <- make.names(rowData(reads)$mgi_symbol, unique = TRUE)
rownames(reads) <- gsub("NA", "ERCC", rownames(reads))
```

# Cell quality control
## Total counts per cell
```{r, fig.height=4.5, fig.width=6}
hist(
  reads$total_counts,
  breaks = 50,
  main = "Reads per cell",
  xlab = "Total Reads"
)
abline(v = 1e5, col = "red")

filter_by_total_counts <- (reads$total_counts > 1e5)

knitr::kable(
  as.data.frame(table(filter_by_total_counts)),
  booktabs = TRUE,
  row.names = FALSE,
  caption = 'The number of cells removed by total counts filter (FALSE)'
)
```


## Detected genes per cell
```{r, fig.height=4.5, fig.width=6}
hist(
  reads$total_features,
  breaks = 50,
  main = "Detected genes per cell",
  xlab = "Total genes"
)
abline(v = c(1000, 6000), col = "red")

filter_by_expr_features <- (reads$total_features > 1000 & reads$total_features < 6000)

knitr::kable(
  as.data.frame(table(filter_by_expr_features)),
  booktabs = TRUE,
  row.names = FALSE,
  caption = 'The number of cells removed by total features filter (FALSE)'
)
```

## mtDNA
```{r, fig.height=4.5, fig.width=6}
plotPhenoData(
  reads,
  aes_string(x = "total_features",
             y = "pct_counts_MT",
             colour = "batch", alpha = 0.8, size = 1), theme = 12) +
  geom_hline(yintercept = 10, col = "red") +
  labs(x = "Total genes", y = "Percentage mitochondrial genes")

filter_by_MT <- reads$pct_counts_MT < 10
knitr::kable(
  as.data.frame(table(filter_by_MT)),
  booktabs = TRUE,
  row.names = FALSE,
  caption = 'The number of cells removed by MT filter (FALSE)'
)
```


## ERCC
```{r}
plotPhenoData(
  reads,
  aes_string(x = "total_features",
             y = "pct_counts_ERCC",
             colour = "cell_type", alpha = 0.9, size = 1), theme = 12) +
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3) + 
  geom_hline(yintercept = c(1, 20), col = "red") +
  labs(x = "Total genes", y = "Percentage ERCC") 

filter_by_ERCC <- (reads$pct_counts_ERCC < 20 & reads$pct_counts_ERCC > 1)

knitr::kable(
  as.data.frame(table(filter_by_ERCC)),
  booktabs = TRUE,
  row.names = FALSE,
  caption = 'The number of cells removed by ERCC filter (FALSE)'
)
```


# Cell filtering
## Manual filter
```{r}
reads$use <- (
  # sufficient features (genes)
  filter_by_expr_features &
    # sufficient molecules counted
    filter_by_total_counts &
    # sufficient endogenous RNA
    filter_by_ERCC &
    # remove cells with unusual number of reads in MT genes
    filter_by_MT
)
table(reads$use)
```

## Automatic filter
```{r}
reads <- plotPCA(
    reads,
    size_by = "total_features", 
    shape_by = "use",
    pca_data_input = "pdata",
    detect_outliers = TRUE,
    return_SCE = TRUE
)
table(reads$outlier)
```

# Gene analysis
## Gene expression
```{r, fig.height=8, fig.width=5}
plotQC(reads, type = "highest-expression")
```

## Gene filtering
```{r}
filter_genes <- apply(counts(reads[, colData(reads)$use]), 1, 
                      function(x) length(x[x > 1]) >= 2)
rowData(reads)$use <- filter_genes

knitr::kable(
  as.data.frame(table(filter_genes)),
  booktabs = TRUE,
  row.names = FALSE,
  caption = 'The number of genes removed by gene filter (FALSE)'
)
```

# Save the data
## Genes and cells after QC
```{r}
dim(reads[rowData(reads)$use, colData(reads)$use])
```

## Calculate raw logcounts
```{r, fig.width=5.5, fig.height=4}
assay(reads, "logcounts_raw") <- log2(counts(reads) + 1)
reducedDim(reads) <- NULL
```

## Remove bad quality libraries
```{r}
reads_qc <- reads[rowData(reads)$use, colData(reads)$use]
```

## Save to RDS file
```{r}
saveRDS(reads, file ="data/1moWTB_reads.rds")
saveRDS(reads_qc, file = "data/1moWTB_reads_qc.rds")
```

# Visualization
PCA and tSNE of cells before and after QC.
```{r}
reads_qc <- reads[rowData(reads)$use, colData(reads)$use]
endog_genes <- !rowData(reads_qc)$is_feature_control
```

## PCA
```{r, fig.width=11, fig.height=4}
pca1 <- plotPCA(reads[endog_genes, ],
                colour_by = "batch",
                size_by = "total_features",
                shape_by = "individual",
                exprs_values = "logcounts_raw",
                theme=12) #+
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
pca2 <- plotPCA(reads_qc[endog_genes, ],
                colour_by = "batch",
                size_by = "total_features",
                shape_by = "individual",
                exprs_values = "logcounts_raw",
                theme=12) #+
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
multiplot(pca1, pca2, cols = 2)
```


## tSNE
```{r, fig.width=14, fig.height=4}
tsne1 <- plotTSNE(reads[endog_genes, ],
                 perplexity = 2,
                 colour_by = "batch",
                 size_by = "total_features",
                 shape_by = "individual",
                 exprs_values = "logcounts_raw",
                 rand_seed = 123456) #+
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
tsne2 <- plotTSNE(reads_qc[endog_genes, ],
                 perplexity = 2,
                 colour_by = "batch",
                 size_by = "total_features",
                 shape_by = "individual",
                 exprs_values = "logcounts_raw",
                 rand_seed = 123456) #+
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
tsne3 <- plotTSNE(reads_qc[endog_genes, ],
                 perplexity = 20,
                 colour_by = "batch",
                 size_by = "total_features",
                 shape_by = "individual",
                 exprs_values = "logcounts_raw",
                 rand_seed = 123456) #+
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
multiplot(tsne1, tsne2, tsne3, cols = 3)
```

