---
title: "Normalization"
author: "Jordi Camps"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
library(scRNA.seq.funcs)
library(SingleCellExperiment)
library(scater)
library(scran)
options(stringsAsFactors = FALSE)
set.seed(1234567)
reads_qc <- readRDS("data/6moWTB_reads.rds")
endog_genes <- !rowData(reads_qc)$is_feature_control
```


## Unnormalized PCA plot
```{r}
plotPCA(
  reads_qc[endog_genes, ],
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual",
  exprs_values = "log2_counts"
) #+
  #scale_colour_brewer(name = "batch", type= "qual", palette = 3)
```

## CPM
### Calculate CPM
```{r}
#assay(reads_qc, "norm_counts") <- log2(calculateCPM(reads_qc, use.size.factors = FALSE) + 1)
logcounts(reads_qc) <- log2(calculateCPM(reads_qc, use.size.factors = FALSE) + 1)
```

### Plot PCA
```{r}
plotPCA(
  reads_qc[endog_genes, ],
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual"
) #+ scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
```

### Plot RLE
```{r}
plotRLE(
  reads_qc[endog_genes, ], 
  exprs_mats = list(Raw = "log2_counts", CPM = "logcounts"),
  exprs_logged = c(TRUE, TRUE),
  colour_by = "batch",
  #style = "full",
  #outlier.alpha = .1,
  #outlier.shape = 3,
  #outlier.size = 0
) #+ 
  #scale_colour_brewer(name = "batch", type= "qual", palette = 3) +
  #scale_fill_brewer(name = "batch", type= "qual", palette = 3)
```


## TMM
### Calculate TMM
```{r}
reads_qc <- normaliseExprs(
  reads_qc,
  method = "TMM",
  feature_set = endog_genes
)
```

### Plot PCA
```{r}
plotPCA(
  reads_qc[endog_genes, ],
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual",
  exprs_values = "norm_cpm"
) #+
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
```

### Plot RLE
```{r}
plotRLE(
  reads_qc[endog_genes, ], 
  exprs_mats = list(Raw = "log2_counts", TMM = "norm_cpm"),
  exprs_logged = c(TRUE, TRUE),
  colour_by = "batch",
  #style = "full",
  outlier.alpha = .1,
  outlier.shape = 3,
  outlier.size = 0
) + 
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) +
  scale_fill_brewer(name = "Batch", type= "qual", palette = 3)
```


## Scran
### Compute size factors for genes and spike-ins
```{r}
#set_exprs(reads_qc, "log2_counts") <- exprs(reads_qc)
qclust <- quickCluster(reads_qc, min.size = 35)
reads_qc <- computeSumFactors(reads_qc, clusters = qclust)
reads_qc <- computeSpikeFactors(reads_qc, general.use = FALSE)
summary(sizeFactors(reads_qc))
reads_qc <- normalize(reads_qc)
```

### Plot PCA
```{r}
plotPCA(
  reads_qc[endog_genes, ],
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual",
  exprs_values = "logcounts",
  theme = 12
) #+
  #scale_colour_brewer(name = "Batch", type= "qual", palette = 3)
```

### Plot RLE
```{r}
plotRLE(
  reads_qc[endog_genes, ], 
  exprs_mats = list(Raw = "log2_counts", Scran = "logcounts"),
  exprs_logged = c(TRUE, TRUE),
  colour_by = "batch",
  #style = "full",
  outlier.alpha = .1,
  outlier.shape = 3,
  outlier.size = 0
) +
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) +
  scale_fill_brewer(name = "Batch", type= "qual", palette = 3) 
```


## DE-Seq2 size factor
### Calculate SF
```{r}
reads_qc <- normaliseExprs(
  reads_qc,
  method = "RLE", 
  feature_set = endog_genes,
  return_log = TRUE,
  return_norm_as_exprs = TRUE
)
```

### Plot PCA
```{r}
plotPCA(
  reads_qc[endog_genes, ],
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual"
) #+ scale_colour_brewer(name = "Batch", type= "qual", palette = 3) 
```

### Plot RLE
```{r}
plotRLE(
  reads_qc[endog_genes, ], 
  exprs_mats = list(Raw = "log2_counts", SF  = "logcounts"),
  exprs_logged = c(TRUE, TRUE),
  colour_by = "batch",
  #style = "full",
  #outlier.alpha = .1,
  #outlier.shape = 3,
  #outlier.size = 0
) #+ scale_colour_brewer(name = "Batch", type= "qual", palette = 3) + 
  #scale_fill_brewer(name = "Batch", type = "qual", palette = 3)
```

## Upperquantile
### Calculate upperquantile
```{r}
reads_qc <- normaliseExprs(
  reads_qc,
  method = "upperquartile", 
  feature_set = endog_genes,
  p = 0.99
)
```

### Plot PCA
```{r}
plotPCA(
  reads_qc[endog_genes, ],
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual",
  exprs_values = "norm_cpm"
) + 
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) 
```

### Plot RLE
```{r}
plotRLE(
  reads_qc[endog_genes, ], 
  exprs_mats = list(Raw = "log2_counts", UQ = "norm_cpm"),
  exprs_logged = c(TRUE, TRUE),
  colour_by = "batch",
  #style = "full",
  outlier.alpha = .1,
  outlier.shape = 3,
  outlier.size = 0
) + 
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) +
  scale_fill_brewer(name = "Batch", type= "qual", palette = 3) 
```


## Downsampling
### Calculate downsampling
```{r}
assay(reads_qc, "norm_counts") <- log2(Down_Sample_Matrix(counts(reads_qc)) + 1)
```

### Plot PCA
```{r}
plotPCA(
  reads_qc[endog_genes, ],
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual",
  exprs_values = "norm_counts"
) + 
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) 
```

### Plot RLE
```{r}
plotRLE(
  reads_qc[endog_genes, ], 
  exprs_mats = list(Raw = "log2_counts", Downsampling = "norm_counts"),
  exprs_logged = c(TRUE, TRUE),
  colour_by = "batch",
  #style = "full",
  outlier.alpha = .1,
  outlier.shape = 3,
  outlier.size = 0
) + 
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) +
  scale_fill_brewer(name = "Batch", type= "qual", palette = 3) 
```


## Normalization for gene/transcript length
### Import and calculate transcript length per gene
```{r}
reads_qc_ann <- reads_qc[!is.na(rowData(reads_qc)$ensembl_gene_id), ]
eff_length <- 
  abs(rowData(reads_qc_ann)$transcript_end - rowData(reads_qc_ann)$transcript_start) / 1000
plot(eff_length, rowMeans(counts(reads_qc_ann)))
tpm(reads_qc_ann) <- log2(calculateTPM(reads_qc_ann, eff_length) + 1)
fpkm(reads_qc_ann) <- log2(calculateFPKM(reads_qc_ann, eff_length) + 1)
```

### Plot PCA
#### FPKM
```{r}
plotPCA(
  reads_qc_ann,
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual",
  exprs_values = "fpkm"
) + 
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) 
```

#### TPM
```{r}
plotPCA(
  reads_qc_ann,
  colour_by = "batch",
  size_by = "total_features",
  shape_by = "individual",
  exprs_values = "tpm"
) + 
  scale_colour_brewer(name = "Batch", type= "qual", palette = 3) +
  scale_fill_brewer(name = "Batch", type= "qual", palette = 3) 
```
