---
title: "PCA and tSNE"
author: "Jordi Camps"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: html_document
---

## Load packages and data
```{r}
library(scTools)
library(dplyr)
library(scater)
options(stringsAsFactors = FALSE)
```

## Load data
```{r}
reads_qc <- readRDS("data/6moWTB_reads_qc.rds")
endog_genes <- !rowData(reads_qc)$is_feature_control
erccs <- rowData(reads_qc)$is_feature_control
```

## Normalize
```{r}
qclust <- quickCluster(reads_qc, min.size = 20)
reads_qc <- computeSumFactors(reads_qc, clusters = qclust)
reads_qc <- computeSpikeFactors(reads_qc, general.use = FALSE)
summary(sizeFactors(reads_qc))
sce <- normalize(reads_qc)
```

## Extract principal components
```{r fig.height=10, fig.width=10}
(sce=plotPCA(sce, exprs_values = "logcounts", ntop=500, colour_by = "cell_type",  ncomponents=4, theme_size =10, return_SCE = TRUE))
sce$PC1 <- reducedDim(sce)[,1]
sce$PC2 <- reducedDim(sce)[,2]
sce$PC3 <- reducedDim(sce)[,3]
sce$PC4 <- reducedDim(sce)[,4]
```

## Extract tSNE dimensions
```{r}
set.seed(123)
(sce <- plotTSNE(sce, colour_by = "sc3_9_clusters", return_SCE = TRUE, perplexity = 20, exprs_values = "logcounts"))
sce$Dim1 <- reducedDims(sce)$TSNE[,1]
sce$Dim2 <- reducedDims(sce)$TSNE[,2]
```

## Plot gene expression on dimensions
```{r}
genes <- c("Dcn", "Sox10", "Plp1", "Myl9")
plot_dims(sce, x = "Dim1", y = "Dim2", var = genes)
```

## Principal component gene weights

Extract genes from different principal components to check for biological processes that are defined in principal components. Done in two steps: 
1. Extracting gene weights from sce_object
2. Saving weights with genes in ascending order per principal component

Extract gene weight for all principal components
500 is the amount of genes we selecte to do the PCA analysis on
```{r}
rv <- rowVars(assay(sce))
selected_genes <- order(rv, decreasing = TRUE)[seq_len(min(500, length(rv)))]#500 is the amount of highest variable genes to include
pca <- prcomp(t(assay(sce)[selected_genes,]))
pca_loadings <- as.data.frame(pca$rotation)
pca_loadings[order(pca_loadings[,1], decreasing = TRUE),]
write.csv(pca_loadings, filename)#fill in filename
```
